Region-network sparsity priors for\\ high-dimensional inference in fMRI data

### 2015-11-26 paper/paper.tex
rnratio slides add

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amsfonts,amsthm}
\usepackage{bbm}
\usepackage{algorithm,algorithmic}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{bbm}
\usepackage[titletoc]{appendix}
\usepackage{wrapfig}
\usepackage{afterpage}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{ulem}
\usepackage{multirow}

@@

\title{Region-network sparsity priors for high-dimensional inference in fMRI data}

@@

\author{Danilo Bzdok, Karl Zilles, Michael Eickenberg,
  Ga\"el Varoquaux, Bertrand Thirion
  INRIA, Parietal team, Saclay, France\\
  CEA, Neurospin, Gif-sur-Yvette, France\\
  firstname.lastname@inria.fr}

@@

\begin{abstract}
Imaging neuroscience links human behavior to aspects of brain
biology in ever-increasing datasets.

@@

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent passage from the classical regime (i.e., "long data")  to
the high-dimensional regime (i.e., "wide data")
\citep{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations available in a data sample.
This scenario challenges many statistical estimators from
classical statistics.
For instance,
linear regression models without additional assumptions
yield an infinity of possible coefficients
and, thus, no solution.
%
\textit{Sparsity} assumptions have consequently been
introduced and made many ill-posed estimation problems tractable
\cite{buhlmann2011statistics, hastie2015statistical}.
Domain-specific structural knowledge can be imposed on the 
statistical estimation by preassuming that not all variables will be
equally important
\citep{bach2012optimization}.
Sparsified supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Yet, what is the most pertinent neurobiological structure
that can successfully tackle the
\textit{curse of dimensionality} in
functional neuroimaging research?

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions probably solve distinct classes of computational processes
\citep{kanwisher2010functional}.
Functional integration, in turn, emphasizes that brain function
is probably enabled by complex connections between these
distinct brain regions.
%
This notion was predominantly derived from
invasive examination of \textit{structure} (i.e., histological preparation),
\textit{connectivity}, (i.e., axonal tracing),
and \textit{functional properties}
(i.e., single cell recordings) in same animals.
Regarding functional segregation into specialized regions,
early histological investigations into the microscopic heterogeneity of
the human cerebral cortex have resulted
in several detailed anatomical maps
\citep{{brodmann1909vergleichende, vogt1919allgemeine}.
Regarding axonal connections,
each such cortical area has been observed
to possess a unique set of incoming and outgoing connections
\citep{passingham2002, young93monkey, scannell95cat}.
%
In particular,
the realization of computational processes (i.e., function)
is probably a consequence of both
local cyto- and chemoarchitectonic infrastructure
and its unique global connectivity profile.
While
cortical modules versus connections between those
reflect 
functional specialization versus functional integration,
\citep{friston2002beyond},
these architectural principles are conceptually inextricable in
the realization of mental operations
\citep{tononi1998complexity}.

method -> type of discovery

SPECIAL
Functional specialization has been
explored with many different research methods.
%
Single cell recordings and microscopic examination
revealed, for instance, the
specialization in the visual cortex into V1, V2, V3, V3A, and V4
\citep{hubel1962receptive, zeki1978functional}.
Tissue lesion reports of the mid-fusiform gyrus of the visual system
frequently indicated impaired
recognition of others' identity from faces
\citep{iaria2008contribution}.
The localization of
sensory, motor, and emotional cognitive functions to cortical areas
in the living brain
has later been enabled by
non-invasive brain imaging with
functional magnetic resonance imaging (fMRI) and
positron emission tomography (PET)
\citep{fristen1997imaging}.
Further,
radioactive mapping the distribution of
neurotransmitter receptors rendered accessible yet another
local characteristic of neuronal populations
\citep{zilles2009receptor}.
% allude to our many-regions layer in the methods
In the computational era,
automatic clustering methods are increasingly employed to
regionally differentiation of the cerebral cortex,
which partly turned out to be more fine-grained than
classical cytoarchitectonic mapping
\citep{behrens03, cbp2015review}.
Today,
high-thoughput computing enables 
ultahigh-resolution 3D models of brain anatomy
at macroscopical to near-cellular scale
\citep{amunts2013bigbrain}.
%
As a crucial common point,
all these methodological classes
yielded neuroscientific findings
that were naturally interpreted
non-overlapping, discrete region compartments
as the basic architectonic framework of cortical organization.

INTEGRAL

Only recently we have shifted the main interpretational focus
from single region to macroscopical networks
in systems neuroscience \citep{yuste2015}.

eemrgent properties emerge from the interaction

In
contrast, functional integration emphasizes that function is an
emergent property of collaborating areas. In fact, the higher the
hierarchical level, the more one cortical area connects to different
large-scale networks, which makes their boundaries less clear
(Yeo/Buckner 2011). The above network-network models naturally embody
functional integration as they capture (partially overlapping)
patterns of spatial coherence.

Systems neuroscience has established existence of a
set of fluctuating yet robust brain networks in humans.

example: DMN network?

Identical neural networks have reappeared across cognitive domains
using diverging methods. These observations prompted widely-adopted
notions, including the ``default-mode network'' \cite{raichle2001} (DMN; Raichle et al.,
2001), ``salience network'' (SN; Seeley et al., 2007), and ``dorsal
attention network'' (DAN; Corbetta and Shulmann,
2002). Developmentally, such large-scale networks emerge during late
fetal growth (Doria et al., 2010), before cognitive capacities mature
in childhood. In adults, nodes of a same cohesive network have more
similar functional profiles than nodes from different networks
(Anderson et al., 2013).
This
suggests network-network relationships as an under-appreciated unit of
functional brain organization.

Onset of a given cognitive task might induce
characteristic changes in functional coupling between large-scale
networks. Indeed, a working-memory task entailed increase in metabolic
activity in DAN regions but decrease in default-mode regions (Fransson
et al., 2006). Notably, the functional connectivity did not change
significantly within either DAN or DMN. Additionally, during auditory
event transitions in another experimental fMRI study, both DAN and SN
increased in activity, whereas the DMN decreased in activity
(Sridharan et al., 2008).

Such mediation between supraordinate networks
has been proposed to involve the right anterior insula (Sridharan et
al., 2008) and right temporo-parietal junction (Bzdok et al.,
2013).
The relevance of networks
possibly extends to psychiatric, neurological, and neurodegenerative
disorders (Seeley et al., 2009; Menon 2011).

=> methods: overlapping, continuous network units
invasive axonal tracing studies,
ICA,
seed-based analyses of BOLD fluctations or
electrophysiological oscillations,
as well as
effective connectivity models
and
graph-theoretical measurements
\citep{friston03dcm, buzsaki2004neuronal,
beckmann2005, bullmore2009complex, jbabdi2013long}.

analysis and integration of information

% study
The present study attempts a combination of specialization and
integration aspects by means of structured sparsity
for high-dimensional inference.
%

adaptive estimator
biologically plausible

recent advances in high-dimensional statistics enable

The present investigation reframed network-network dynamics as a
multivariate statistical learning problem.
capitalized on two independent, large datasets (n=500 and n=80) with
18 typical neuroimaging tasks each, covering a spectrum of humans'
cognitive apparatus.

determined a plausible, quantitative
combination of the brain signal's modes of variation. This allowed
avoiding the curse of dimensionality and making the machine-learning
results human-interpretable by quantitative association with
psychological concepts.

importance of brain regions and networks is probably a matter of
degree.
We estimate that degree from the data

help reintegrate the divorced
interpretational streams based on
region- and network-focused neuroscientific investigations.

a compartmental model of
functional organization in the cerebral cortex can be derived

@@

we need to inject domain knowledge into
statistical estimations to harness the curse of dimensionality.

should be able to estimate voxel level
while taking into account known supravoxel structure.

@@

OUTLOOK:
robust high-dimensional inference will be key
to prediction in single individual,
such as for personalized medicine
\citep{gabrieli2015prediction}

@@

\small
\bibliographystyle{splncs03}
\bibliography{paper_refs}

>>>>

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amsfonts,amsthm}
\usepackage{bbm}
\usepackage{algorithm,algorithmic}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{bbm}
\usepackage[titletoc]{appendix}
\usepackage{wrapfig}
\usepackage{afterpage}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{ulem}
\usepackage{multirow}
\usepackage{comment}

@@

\title{Region-network sparsity priors for\\
high-dimensional inference in fMRI data}

@@

\author{Danilo Bzdok, Karl Zilles, Michael Eickenberg,
  Ga\"el Varoquaux, Bertrand Thirion\\
  Department of Psychiatry, Psychotherapy and Psychosomatics, RWTH Aachen, Germany\\
  INRIA, Parietal team, Saclay, France\\
  CEA, Neurospin, Gif-sur-Yvette, France\\
  firstname.lastname@inria.fr}

@@

\begin{abstract}
% Imaging neuroscience links human behavior to aspects of brain
% biology in ever-increasing datasets.

@@

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent passage from the classical regime (i.e., "long data")  to
the high-dimensional regime (i.e., "wide data")
\cite{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations available in a data sample.
This scenario challenges many statistical estimators from
classical statistics.
For instance,
linear regression models without additional assumptions
yield an infinity of possible coefficients
and, thus, no solution.
%
\textit{Sparsity} assumptions have consequently been
introduced and made many ill-posed estimation problems tractable
\cite{buhlmann2011statistics, hastie2015statistical}.
Domain-specific structure can be imposed on the 
statistical estimation,
thus preassuming that variables have unequal importance
and follow existing knowledge
\cite{bach2012optimization}.
Sparsified supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Yet, what neurobiological structure suggests itself
to successfully tackle the
\textit{curse of dimensionality} in
functional neuroimaging research?

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions probably solve distinct classes of computational processes
\cite{kanwisher2010functional}.
Functional integration, in turn, emphasizes that brain function
is probably enabled by complex connections between these
distinct brain regions \cite{sporns14nn}.
%
These notions were predominantly derived from
invasive examination of \textit{structure} (i.e., histological preparation),
\textit{connectivity}, (i.e., axonal tracing),
and \textit{functional properties}
(i.e., single cell recordings) in same animals.
Regarding functional segregation into specialized regions,
early histological investigations into the microscopic heterogeneity of
the human cerebral cortex have resulted
in several detailed anatomical maps
\cite{brodmann1909vergleichende, vogt1919allgemeine}.
Regarding axonal connections,
each such cortical area has been observed
to possess a unique set of incoming and outgoing connections
\cite{passingham2002, young93monkey, scannell95cat}.
%
Both
local cyto- and chemoarchitectonic infrastructure
and its unique global connectivity profile
together realize computational processes (i.e., function).
While
cortical modules versus connections between those
reflect 
functional specialization versus functional integration
\cite{friston2002beyond, mesulam_sensation},
both architectural principles are conceptually inextricable when
explaining the emergence of mental operations
\cite{tononi1998complexity}.

% SPECIAL
Functional specialization has been
explored and interpreted based on many different research methods.
%
Single cell recordings and microscopic examination
revealed, for instance, the
specialization in the visual cortex into V1, V2, V3, V3A, and V4
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system
were frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
The localization of
sensory, motor, and emotional cognitive functions to cortical areas
in the living brain
has later been enabled by
non-invasive brain imaging with
functional magnetic resonance imaging (fMRI) and
positron emission tomography (PET)
\cite{fristen1997imaging}.
Further,
radioactive mapping of neurotransmitter receptor distributions
rendered accessible yet another
local characteristic of neuronal populations
\cite{zilles2009receptor}.
% allude to our many-regions layer in our methods
In the computational era,
automatic clustering methods are increasingly employed to
regionally differentiate the cerebral cortex,
which can partly be more fine-grained than
classical cytoarchitectonic borders
\cite{behrens03, cbp2015review}.
Today,
high-thoughput computing enables 
ultahigh-resolution 3D models of brain anatomy
at macroscopical to near-cellular scale
\cite{amunts2013bigbrain}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architectonic framework of brain organization.

% INTEGRAL
It is more recent
that the main interpretational focus has shifted
from circumscribed regions to network stratifications
in systems neuroscience \cite{yuste2015, stephan_dys}.
The higher the processing hierarchy level,
the more a cortical area is known to
connect to different large-scale networks
\cite{yeo11}.
%
Invasive axonal tracing studies in monkeys were complemented
by in-vivo diffusion MRI tractography in humans
as a now frequently employed method to
outline fiber bundles between brain regions
\cite{jbabdi2013long}.
Besides analyses of
electrophysiological oscillations
\cite{buzsaki2004neuronal}
and
graph-theoretical properties \cite{bullmore2009complex},
studies of
functional connectivity \cite{buckner2013opportunities} and
independent component analysis (ICA) \cite{beckmann2005}
became the workhorses of network discovery
in functional neuroimaging.
These revealed the important implication of
canonical brain networks across cognitive domains,
including the so-called
``default-mode network'' \cite{raichle2001pnas},
``salience network'' \cite{seeley2007dissociable},
and ``dorsal attention network'' \cite{corbettashul2008}. 
Characteristic changes in the configuration of
these large-scale networks
were repeatedly observed to be induced
by the onset of given cognitive tasks \cite{fransson2006}.
Such task-induced mechanisms orchastrating supraordinate networks
might be subserved by the right anterior insula \cite{sridh2008}
and temporo-parietal junction \cite{bzdok2013tpj}.
%
Ultimately,
interpretation of findings from all these methods naturally embraces
cross-regional integration by
overlapping network compartments
as the basic architectonic framework of brain organization,
in stark constrast to methods examining regional specialization.

% study
Building on these interpretational traditions,
the present study proposes to incorporate
known architectonic aspects from both
anatomical segregation and integration
into unified classification and regression algorithms
by means  of structured sparsity assumptions.
The "true" relative importance of 
local region compartments and global network compartments
is typically unknown
but
probably varying in degree
across diverse neuroscientific questions.
Recent advances in multivariate statistical learning techniques
enable such adaptive estimators with
neurobiologically plausible region and network priors.
Using a large reference dataset,
we demonstrated that domain-informed supervised models
gracefully handle the curse of dimensionality,
yield many more human-interpretable results,
and generalize better to new samples
than domain-na\"ive models.

\begin{comment}

@@

we need to inject domain knowledge into
statistical estimations to harness the curse of dimensionality.
two neurobiological design principles

integrative processes

should be able to estimate voxel level
while taking into account known supravoxel structure.
is instrumental in
Developmentally, such large-scale networks emerge during late
fetal growth (Doria et al., 2010), before cognitive capacities mature
in childhood.

In adults, nodes of a same cohesive network have more
similar functional profiles than nodes from different networks
(Anderson et al., 2013).

neurobiologically motivated restrictions to complexity circumvented
the curse of dimensionality

@@

OUTLOOK:
robust high-dimensional inference will be key
to prediction in single individual,
such as for personalized medicine
\cite{gabrieli2015prediction}

@@

\end{comment}

\small
% \bibliographystyle{plainnat}
\bibliographystyle{splncs03}
\bibliography{paper_refs}
